### 2020-12-05 paper/2020-10-revision-background.tex
Shorten version of conceptual framework

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I argue that female education and area of residence are the most important factors 
to consider in the empirical analysis and discuss how they tie into the three theories.

% Economic conditions

The dramatic improvements in economic conditions in India are best illustrated by the 
doubling of real wages between 1987 and 2011 \citep{Klasen2015}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as a good becomes more expensive, people substitute 
away from that good and towards other goods.
With higher wages, this means that people work more and spend less time on non-wage 
earning activities, such as children or leisure.
The income effect captures that, holding hours worked constant, higher wages increase the 
available income. 
This positive income effect leads people to spend more time on time-intensive activities 
such as children and leisure and less time working.

Empirically, higher female wages generally lowers fertility, whereas higher male 
wages increases fertility \citet{Hotz1997,schultz97}.
This pattern is consistent with women spending substantially more time on child care 
than men, leading to the substitution effect dominating for women whereas the income
effect dominates for men.

In addition to lowering fertility, higher female wages will tend to shorten birth spacing 
if having children requires the mother to reduce her economic activities.
Specifically, shortening birth spacing allows parents to take advantage of economies of 
scale in childrearing---for example, looking after two children simultaneously requires 
less than double the time needed to look after one child \citep{Vijverberg1982,Hotz1997}.
The reduction in the length of birth intervals from higher female wages will be more 
pronounced the more women participate in the labor force rather than working at home as 
economic activities at home are more compatible with child rearing.

% Child outcomes
Parents care not only about how many children they have but also about the outcomes
for the children, and, everything else equal, more children means fewer resources to 
invest per child.
Hence, increasing returns to education provides parents with an incentive to reduce 
fertility and invest more in each of the children's education \citep{Rosenzweig1982a}.
Furthermore, lower expected mortality means that fewer births are required to reach a 
desired number of surviving children and allow parents to invest more in each child.

Longer spacing between births is generally expected to have positive effects on 
child outcomes and if fertility falls because of increased desired for better child
outcomes we would also expect spacing to increase.
The clearest example of birth spacing affecting child outcomes is for health,
where very short spacing---approximately 24 months or less---leads to worse health 
and mortality outcomes for children 
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012}.
Whether longer spacing results in increased parental attention per child and better 
human capital outcome is unclear as the evidence is mixed for developed countries and 
nonexistent for developing countries
\citep{Zajonc1975,Zajonc1976,Razin1980,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, son preference has only a minor effect on average fertility, but
differential stopping behavior does affect the sex composition and fertility of
individual families \citet{repetto72,leung94,clark00,Basu2010,Barcellos2014}. 
Furthermore, as mentioned above, stronger son preference leads to shorter spacing and  
worse health outcomes for daughters when there are no sons in the household 
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.

The use of sex selection allows parents to avoid giving birth to a girl but each
abortion increases the expected interval to the next birth by 6--12 months.
This increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation.

Holding son preference constant, lower desired fertility increases the use of sex 
selection \citep{Portner2015b,Jayachandran2017}.  
Furthermore, the higher the parity the more likely the use of sex selection if the 
desired number of sons has not been reached.

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest 
available proxies wages and the access to and cost of sex selection, none of which are 
available in the data.
In addition, earlier research show that both higher female education and urbanization are 
associated with lower fertility and increased use of sex selection, and, therefore,
likely also changes in birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas, as illustrated by Figure 
\ref{fig:education_over_time}, which shows the distribution of schooling by birth 
cohort for all urban and rural women, 20 years or older, using the four rounds of the NFHS.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older at survey}
\label{fig:education_over_time}
\end{figure}

There are three reasons that families in urban areas are likely to have fewer children and 
higher rates of sex selection. 
First, the higher cost of children in urban than rural areas lowers fertility and 
increases the use of sex selection.
Second, better access to health care in urban areas lowers the mortality risk, further 
reducing fertility.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
Finally, the better access to health care also imply easier access to sex selection and 
the higher income of urban households makes it more affordable.

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but the low and declining female labor 
force participation, especially for younger women, suggests that families face little 
incentive to space children more closely together for economic reasons.
Even as the level of female education increased, the female labor force participation 
in both urban and rural areas stagnated or decreased 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.

Female labor force participation does, however, still depend on education.
In line with other countries, the NFHS data show a U-shaped relationship between 
education and working for married women, with women with either no education or 12 or more 
years the most likely to participate in the labor force and women with 8--11 years of 
education the least \citep{Goldin1994,Chatterjee2018}.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

One explanation for the low female labor force participation, especially for women
with 8--11 years of education, is that household income has increased so substantially 
that the income effect dominates any substitution effect.
Two findings support this interpretation.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to their own wages \citep{Bhargava2018}.

Furthermore, female education undoubtedly represents more than the value of women's time.
Beside propagation of low fertility norms, two examples are the productions of child 
health and offspring human capital.

First, children of more educated women have better health outcomes and lower mortality 
than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Especially of interest here is that, although short birth spacing generally leads to worse 
health outcomes, the more educated the mother, the more muted this relationship is, which, 
at least partly, removes the health outcome incentive for longer birth spacing
\citep{Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
Hence, the increasing return to male education may lead to an increased demand for 
better-educated mothers, which, combined with lower fertility and an increased use
of sex selection when available, allow families to achieve higher quality sons.
In this explanation, the associations between higher female education, lower fertility, 
and more sex selection are not a causal effect driven by higher female wages but instead 
the result of the increased return to male education.

% Sanskritization
Although the expansion in female education might bring in groups with initially different 
norms, the process of ``Sanskritization'' implies that as lower-castes females gain access 
to education, they adopt higher-caste norms such as stronger son preference and a retraction 
from the formal labor market \citep{Srinivas1956}.
The low and declining female labor force participation as husbands' wages increase 
is one indication that this process still operates \citep{Abraham2013,Chatterjee2018}.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.

>>>>

This section provides a conceptual framework for understanding how birth spacing changes
and guiding the subsequent analysis.
I first introduce three potential explanations that link fertility and birth 
spacing decisions: economic conditions, desired child outcomes, and son preference 
\citep{Casterline2016,Portner2018}.
Next, I discuss why female education and area of residence are the principle factors 
in the empirical analyses and how they tie into the three explanations.

% Economic conditions
The improvements in economic condition, especially the doubling of real wages,
are likely to affect fertility although the direction is ambiguous, as, empirically, 
higher female wages lowers fertility, whereas higher male wages increases fertility 
\citep{Hotz1997,schultz97}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as wages increase time becomes more expensive, and
people therefore work more and spend less time on non-wage earning activities, such as 
children or leisure.
The income effect captures that higher wages increase the available income, which
leads people to spend more time on time-intensive activities and less time working.
Since women spend substantially more time on childrearing than men, the substitution effect 
dominates for women but the income effect for men.

If having children requires the mother to curtail her economic activities, higher
female wages, in addition to lowering fertility, will tend to shorten birth spacing.
Shortening birth spacing allows parents to take advantage of economies of scale in 
childrearing---for example, looking after two children simultaneously requires 
less than double the time needed to look after one child \citep{Vijverberg1982,Hotz1997}.

% Child outcomes
Parents care both about how many children they have and their outcomes, and with 
increasing returns to education and lower offspring mortality, parents are likely to 
reduce fertility and invest more in each child \citep{Rosenzweig1982a,Wolpin1997}.
The higher return to education means that parents have more of an incentive to 
invest in their children's education, which increases the cost of children and 
lowers fertility. 
Lower expected mortality means that fewer births are required to reach a 
desired number of surviving children and allow parents to invest more in each child.

Longer spacing between births is generally expected to have positive effects on 
child outcomes and if fertility falls because of increased desired for better child
outcomes we would also expect spacing to increase.
The clearest example of birth spacing affecting child outcomes is for health,
where very short spacing---approximately 24 months or less---leads to worse health 
and mortality outcomes for children 
\citep{Whitworth2002,Conde-Agudelo2012}.
Whether longer spacing also results in better human capital outcome is unclear as the 
evidence is mixed for developed countries and nonexistent for developing countries
\citep{Zajonc1976,Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

% Son preference 
Empirically, a stronger son preference has only a minor effect on average fertility, but
son preference still affects the sex composition and fertility across families through 
differential stopping behavior and leads to shorter spacing and  worse health outcomes for 
daughters when there are no sons in the household 
\citep{repetto72,clark00,Whitworth2002,Basu2010,Jayachandran2011,Barcellos2014}.

Another result of son preference is the use of sex selection, which enables parents to 
avoid giving birth to a girl but increases the expected interval to the next birth
by 6--12 months for each abortion.
This increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation.

The use of sex selection, and therefore the length of birth intervals, increases with
lower desired fertility and the higher the parity provided the desired number of sons has 
not been reached \citep{Portner2015b,Jayachandran2017}.

% Factors/predictions
Although the explanations above are based on economic conditions, desired child outcome,
son preference, and the access to sex selection, none of these are available in the
data, so I instead use female education and area of residence as the two main explanatory 
variables in the empirical analyses.
Female education and area of residence are the closest available proxies and earlier 
research show that both higher female education and urbanization are associated with 
lower fertility and increased use of sex selection, and, therefore, likely also changes 
in birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Furthermore, both variables are interesting given the tremendous increases in 
both education and urbanization in India; 
Figure \ref{fig:education_over_time} show the substantial increase in female education
in both rural and urban areas, although it is still considerably higher in urban than in 
rural areas and the proportion of the population living in urban areas almost doubled from 
18\% in 1960 to 35\% in 2019 \citep{United-Nations2019}.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older
using the four rounds of the NFHS}
\label{fig:education_over_time}
\end{figure}

Women living in urban areas are likely to have lower fertility and higher use
of sex selection than women in rural areas.
The lower fertility is the result of higher cost of children and lower child mortality 
risk in urban compared to rural areas.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
The higher use of sex selection comes partly from the lower fertility and partly from 
women in urban areas having higher income and better access to health care facilities.

As women receive more education their potential wages increase, which should shorten 
birth spacing and lower fertility, but the low and declining female labor force 
participation in India, especially for younger women, suggests that families face little 
incentive to space children more closely together for economic reasons.
Even as women's education increased, their labor force participation stagnated or 
decreased and is now one of the lowest outside the Middle East 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018}.%
\footnote{
In line with other countries, there is still a U-shaped relationship between education and 
labor force participation for married women, with the lowest participation for 
women with 8--11 years of education \citep{Goldin1994,Chatterjee2018}.
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

One explanation for the low female labor force participation is that male wages have 
increased so substantially that the income effect dominates any substitution effect.
The average male wage is almost 70\% higher than the female wage and women’s labor 
supply appears to be more negatively elastic to husbands' wages than positively elastic 
to their own wages \citep{Klasen2015,Bhargava2018}.

Furthermore, the increasing return to male education may make it appear as if higher 
female education results in lower fertility and more sex selection, if women with more 
education are better at producing child human capital as suggested by \citet{Behrman1999},.
The higher return to male education provides an additional incentive to invest in sons'
education, and one such investment is a better-educated mother.
As discussed above, the higher returns should also lower desired fertility, making the use 
of sex selection even more appealing, and possibly increases birth spacing.

% Sanskritization
Although the expansion in female education might bring in groups with initially different 
norms, the process of ``Sanskritization'' implies that as lower-castes females gain access 
to education, they adopt higher-caste norms such as stronger son preference and a retraction 
from the formal labor market \citep{Srinivas1956}.
The low and declining female labor force participation as husbands' wages increase 
indicates that this process still operates \citep{Abraham2013,Chatterjee2018}.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect longer birth spacing over time, independent of education 
levels.
Furthermore, birth spacing likely increases the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change the 
use of sex selection.
